---
title: "Calculating the Topographic Position Index"
lang: en
---

::: {.callout-note}
[Japanese version](../../notebooks/topographic_position_index.qmd)
:::

The Topographic Position Index (TPI) is an index for quantifying the local position of terrain.
It is calculated as the difference between the elevation of a target cell and the mean elevation of its surrounding cells.
Positive values indicate that the cell is higher than its surroundings, while negative values indicate that it is lower.

$$
TPI = elevation - mean elevation of surrounding cells
$$

Here, we calculate TPI from a DEM (Digital Elevation Model) with the R package [terra](https://rspatial.github.io/terra/index.html).

The example uses sample data from the [`whitebox`](https://whiteboxr.gishub.org/index.html) package.

```{r}
# install.packages("terra") # if terra is not installed yet
# renv::install("terra") # if you use renv
# install.packages("whitebox") # if whitebox is not installed yet
# renv::install("whitebox") # if you use renv
library(terra)
library(whitebox)

dem_path <- sample_dem_data()
dem <- rast(dem_path)
plot(dem)
```

After deciding how many surrounding cells to use for each elevation value, use `terra::focal()` to calculate the mean elevation of the neighboring cells.
Here, we use a 3 x 3 window.

```{r}
w <- matrix(1, nrow = 3, ncol = 3) # 3 x 3 window
w[2, 2] <- NA # exclude the center cell from the neighborhood mean

mean_neigh <- focal(
  dem,
  w = w,
  fun = mean,
  na.rm = TRUE
)

tpi_3_3 <- dem - mean_neigh
plot(tpi_3_3, main = "TPI (3x3)")
```

Interpretation is as follows.

| TPI | Meaning |
| --: | ------- |
| Positive | Higher than the surroundings; closer to a ridge or convex landform |
| Around 0 | Similar to the surroundings |
| Negative | Lower than the surroundings; closer to a valley or concave landform |

The spatial scale used for TPI should be chosen according to the purpose of the analysis.
