---
title: "Calculating the Topographic Wetness Index"
lang: en
---

::: {.callout-note}
[Japanese version](../../notebooks/topographic_wetness_index.qmd)
:::

## Overview

This notebook shows how to calculate the Topographic Wetness Index (TWI) from a DEM (digital elevation model) with the R package [`whitebox`](https://whiteboxr.gishub.org/index.html).

## Workflow

The workflow is as follows.

```{mermaid}
graph TD
    A[Read DEM] --> B[Depression breaching]
    B --> C[Calculate flow accumulation]
    C --> D[Calculate slope]
    D --> E[Calculate TWI]
```

## Setup

If this is your first time using `whitebox`, run the following code to install [WhiteboxTools](https://www.whiteboxgeo.com/).
The calculation itself is performed by the external WhiteboxTools executable; R works as an interface to call it.

```{r whitebox-install, eval=FALSE}
# Run only once if needed
# renv::install("terra")
# renv::install("whitebox")
whitebox::install_whitebox()
```

Load the libraries.
The [`terra`](https://rspatial.github.io/terra/index.html) package is used for raster data handling.
The [`whitebox`](https://whiteboxr.gishub.org/index.html) package is used to call WhiteboxTools from R.

```{r}
library(terra)
library(whitebox)

# Get the WhiteboxTools executable path and confirm that it exists
wbt_path <- wbt_exe_path(shell_quote = FALSE)
if (!whitebox::check_whitebox_binary()) {
  stop(
    "WhiteboxTools executable was not found: ",
    wbt_path,
    "\n",
    "Run whitebox::install_whitebox() once, then rerun this notebook.",
    call. = FALSE
  )
}

# Check the WhiteboxTools path and version
wbt_path
whitebox::wbt_version()

# Create the output directory
output_dir <- "output"
dir.create(output_dir, showWarnings = FALSE)
```

## Read and visualize the DEM

This example uses sample data from the `whitebox` package.

```{r}
dem_path <- sample_dem_data()
dem <- rast(dem_path)
print(dem)

# Visualization
plot(dem, axes = FALSE, main = "DEM (Original)")
```

::: {.callout-important}

## Coordinate Reference System

The DEM must use a projected coordinate reference system.
If it uses a geographic coordinate reference system, distance and area calculations are not accurate, and the resulting TWI may be unreliable.

The following code checks whether the CRS is set and whether it is projected.

```{r}
# Check the coordinate reference system
if (!nzchar(crs(dem))) {
  stop("DEM has no coordinate reference system.", call. = FALSE)
}
# Check whether the DEM uses a projected coordinate reference system
if (isTRUE(is.lonlat(dem))) {
  stop("DEM must use a projected coordinate reference system.", call. = FALSE)
}
```

:::

## Depression breaching

First, run depression breaching to remove depressions from the DEM.
This is an important preprocessing step for hydrological terrain analysis.

The `wbt_breach_depressions_least_cost()` function performs depression breaching.
The required arguments are:

- `dem`: file path to the input DEM
- `output`: file path for the output raster
- `dist`: search distance for removing depressions; the unit depends on the DEM coordinate system

The `fill` argument is `TRUE` by default, which fills unresolved depressions after breaching.
In most cases, leaving `fill = TRUE` should be fine.

```{r}
wbt_breach_depressions_least_cost(
  dem = dem_path,
  output = file.path(output_dir, "dem_breached.tif"),
  dist = 20,
  fill = TRUE
)
dem_breached <- rast(file.path(output_dir, "dem_breached.tif"))
```

Compare the breached DEM with the original DEM.

```{r}
# Compare the breached DEM with the original DEM
local({
  old_mar <- par("mar")
  on.exit(par(mar = old_mar), add = TRUE)

  par(mar = c(2, 1, 2, 1))
  plot(dem, axes = FALSE, main = "DEM (Original)")
  plot(dem_breached, axes = FALSE, main = "DEM (Breached Depressions)")
  plot(
    dem - dem_breached,
    axes = FALSE,
    main = "Difference (Original - Breached)"
  )
})
```

In this sample data, elevation changes caused by depression breaching were very small.

## Flow accumulation

Next, calculate flow accumulation.
This step estimates how much upslope contributing area flows into each cell.

```{r}
wbt_d8_flow_accumulation(
  input = file.path(output_dir, "dem_breached.tif"),
  output = file.path(output_dir, "d8_accum.tif"),
  out_type = "specific contributing area",
  log = FALSE
)
dem_breached_d8_accum <- rast(file.path(
  output_dir,
  "d8_accum.tif"
))
plot(dem_breached_d8_accum, axes = FALSE, main = "Flow Accumulation (D8)")
```

This calculates flow accumulation with the D8 algorithm.
`out_type = "specific contributing area"` means that the output is specific catchment area, or contributing area per unit contour width.

Other options include `wbt_d_inf_flow_accumulation()` for D-infinity flow accumulation and `wbt_fd8_flow_accumulation()` for FD8 flow accumulation.
I am still learning the details, but the appropriate method appears to depend on the analysis.

## Slope

Next, calculate slope from the DEM with `wbt_slope()`.

```{r}
wbt_slope(
  dem = file.path(output_dir, "dem_breached.tif"),
  output = file.path(output_dir, "d8_slope.tif"),
  units = "degrees"
)
dem_breached_slope <- rast(file.path(
  output_dir,
  "d8_slope.tif"
))
plot(dem_breached_slope, axes = FALSE, main = "Slope (degrees)")
```

## TWI (Topographic Wetness Index)

Use `wbt_wetness_index()` to calculate TWI.
This function takes flow accumulation and slope rasters as inputs and calculates the Topographic Wetness Index.

```{r}
wbt_wetness_index(
  sca = file.path(output_dir, "d8_accum.tif"),
  slope = file.path(output_dir, "d8_slope.tif"),
  output = file.path(output_dir, "d8_twi.tif")
)
twi <- rast(file.path(output_dir, "d8_twi.tif"))

plot(twi, axes = FALSE, main = "TWI (Topographic Wetness Index)")
```
